Equalizer Programming Guide

### 2008-08-22 Developer/ProgrammingGuide/paper.tex
Programming Guide: Partial update for OSWindow changes.
[ ] May break build
[ ] Breaks existing applications
[ ] Bugfix
[ ] New Feature
[ ] Cleanup
[ ] Optimization
[x] Documentation

\date{
  \vspace{2cm}
  \begin{figure}[ht]
    \hspace{-4.5cm}\includegraphics[width=22cm]{images/teaser.pdf}
  \end{figure}
  \vspace{1cm}
%  \textbf{RELEASE CANDIDATE}\\[\medskipamount]
  Version 1.3.6 for Equalizer 0.5.3+svn\\[\medskipamount]
  \today
}

@@

\clearpage
\thispagestyle{empty}
\tableofcontents
\vfill{\center\begin{tabularx}{\textwidth}{|l|l|X|}
    \hline
    \bf Rev & \bf Date     & \bf Changes \\
    \hline
    1.0     & Oct 28, 2007 & Initial Version for Equalizer 0.4\\
    1.2     & Apr 15, 2008 & Version for Equalizer 0.5\\
    1.3.1   & May 02, 2008 & Add \sref{sStatistics}, improved images based on 1.2 print\\
    1.3.2   & May 05, 2008 & Add network distribution on eqPly model data\\
    1.3.3   & May 10-19, 2008 & add \aref{aFileFormat}\\
    1.3.4   & July 3, 2008 & Update \sref{sStatistics}\\
    1.3.5   & July 7, 2008 & Update after namespace naming cleanup\\
    1.3.6   & July 20, 2008 & Add accum and samples to \aref{aFileFormat}, add \sref{sDrawableConfig}\\
    \hline
  \end{tabularx}}
\thispagestyle{empty}
\clearpage

@@

The \textsf{eq::Window} class holds a drawable and an OpenGL
context. The drawable can be an on-screen window or an off-screen
PBuffer. Framebuffer object (FBO) support will be implement by later
version of Equalizer. The window holds window-system-specific handles to
the drawable and context, e.g., an X11 window \textsf{XID} and
\textsf{GLXContext} for the glX window system.

@@

\subsection{Window}

The Equalizer window holds an OpenGL drawable and a rendering
context. When using the default window initialization functions, all
windows of a pipe share the OpenGL context. This allows reuse of OpenGL
objects such as display lists and textures between all windows of one
pipe.

@@

@@

The initialization sequence uses multiple, overrideable task
methods. The main task method \textsf{configInit} executes a `child'
task method to create the drawable and context. The child task method
depends on the window system of the pipe. The default implementations of
\textsf{configInitGLX}, \textsf{configInitWGL} or \textsf{configInitAGL}
create an on-screen window using OS-specific methods. If the OpenGL
drawable and context were created successfully, \textsf{configInit} calls
\textsf{configInitGL}, which performs the generic OpenGL
state initialization. The default implementation sets up some typical
OpenGL state, e.g., it enables the depth test.

\fig{fWindowInit} shows a flow chart of the window initialization. The
colored functions are task methods and can be replaced by
application-specific implementations.

The window-system specific initialization takes into account various
attributes set in the configuration file. Attributes include the size of
the various frame buffer planes  (color, alpha, depth, stencil) as well
as other framebuffer attributes, such as quad-buffered stereo,
doublebuffering, fullscreen mode and window decorations. Some of the
attributes, such as stereo, doublebuffer and stencil can be set to
\textsf{eq::AUTO}, in which case Equalizer will test for their availability and 
enable them if possible.

For the window-system specific initialization, \textsf{eqPly} uses the
default Equalizer implemenation. The \textsf{eqPly} window
initialization only overrides the OpenGL-specific initialization
function in order to initialize a state object and an overlay logo. This
function is only called if an OpenGL context was created and made current:

@@

    eq::Pipe*  pipe        = getPipe();
    Window*    firstWindow = static_cast< Window* >( pipe->getWindows()[0] );

    EQASSERT( !_state );

    if( firstWindow == this )
    {
        _state = new VertexBufferState( glewGetContext( ));

        const Config*   config   = static_cast< const Config* >( getConfig( ));
        const InitData& initData = config->getInitData();

        if( initData.useVBOs() )
        {
            // Check if VBO funcs available, else leave DISPLAY_LIST_MODE on
            if( GLEW_VERSION_1_5 )
            {
                _state->setRenderMode( mesh::BUFFER_OBJECT_MODE );
                EQINFO << "VBO rendering enabled" << endl;
            }
            else
                EQWARN << "VBO function pointers missing, using display lists" 
                       << endl;
        }

        if( initData.useGLSL() )
        {
            // Check if all functions are available
            if( GLEW_VERSION_2_0 )
            {
                EQINFO << "Shaders supported, attempting to load" << endl;
                _loadShaders();
            }
            else
                EQWARN << "Shader function pointers missing, using fixed "
                       << "pipeline" << endl;
        }

        _loadLogo();
    }
    else
    {
        _state       = firstWindow->_state;
        _logoTexture = firstWindow->_logoTexture;
        _logoSize    = firstWindow->_logoSize;
    }

    if( !_state ) // happens if first window failed to initialize
    {
        setErrorMessage( "No state handler object" );
        return false;
    }

    // turn off OpenGL lighting if we are using our own shaders
    if( _state->getProgram( pipe ) != VertexBufferState::FAILED )
        glDisable( GL_LIGHTING );

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/windowInit.pdf}
  {\caption{\small\label{fWindowInit}Window Initialization}}
\end{wrapfigure}
The state object is used to handle the creation of OpenGL objects in a
multipipe, multi-threaded execution environment. It uses an object
manager, which is described in detail in \sref{sObjectManager}. It is
used in conjunction with a reference-counting smart-pointer here, since it is
potentially `owned' by multiple windows at the same time.

@@

The window exit function de-allocates all OpenGL objects when the state
object is about to be disposed. The object manager does not delete the
object in its destructor, since it does not know if an OpenGL context is
still current. Additionally, \textsf{eq::Window::configExit} is called
to destroy the drawable and context:

{\footnotesize\begin{lstlisting}
bool Window::configExit()
{
    if( _state.isValid() && _state->getRefCount() == 1 )
        _state->deleteAll();

    _state = 0;
    return eq::Window::configExit();
}
\end{lstlisting}}

@@

The model data is spatially organized in an 3-dimensional
kD-tree\footnote{See also \link{http://en.wikipedia.org/wiki/Kd-tree}}
for efficient view frustum culling. When the model is loaded by
\textsf{Node::config\-Init}, it is preprocessed into the kD-tree and each
node of the tree gets a database range assigned. The root node
has the range [0, 1], its left child [0, 0.5] and its right child [0.5,
1], and so on for all nodes in the tree. The preprocessed model is saved
in a binary format for accelerating subsequent use.

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/render.pdf}
  {\caption{\small\label{fRender}Main Render Loop}}
\end{wrapfigure}
The rendering loop maintains a list of candidates to render, which
initially contains the root node. Each candidate of this list is tested
for full visiblity against the frustum and range, and rendered if
visible. It is dropped if it is fully invisible or fully out of
range. If it is partially visible or partially in range, the children of
the node are added to the candidate list. \fig{fRender} shows a flow
chart of the rendering algorithm, which performs efficient view frustum
and range culling.

@@

During this initialization, Equalizer creates and manages two threads
for communication, the receiver thread and the command thread. Normally
no application code is executed from these two threads.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/threads.pdf}
  {\caption{\small\label{fThreads}Threads within one node process}}
\end{wrapfigure}
The command thread processes Equalizer-related request from other nodes,
for example during \textsf{eq::net::Object} mapping. In some special cases
the command thread executes application code, for example when a remote
node maps a static or unbuffered object,
\textsf{Object::getInstanceData} is called from the command thread.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/frameSync.pdf}
  {\caption{\small\label{fFrameSync}Per-Node Frame Synchronization}}
\end{wrapfigure}
Each entity uses \textsf{waitFrame\-Started} to block on the parent's
\textsf{startFrame}, e.g., \textsf{Pipe::\-frame\-Start} calls
\textsf{Node::wait\-Frame\-Started} to wait for the corresponding
\textsf{Node::start\-Fra\-me}. This explicit synchronization allows to
update non-critical data before synchronizing with
\textsf{waitFrameStarted}, or after unlocking using
\textsf{start\-Fra\-me}. \fig{fFrameSync} illustrates this
synchronization model.

@@

\fig{fFrameSync} illustrates the local frame synchronization. By
removing the calls to \textsf{waitFrameLocal}, the node thread can
disable the synchronization with the pipe threads, which causes the node
operations to overlap with the rendering.

@@

This section explains window initialization in detail. It discusses in
detail the handling on the different window systems. The entry point for
the window initialization is the task method
\textsf{Window::configInit}. This task method first calls a
window-system-specific task method, and then
\textsf{Window::configInitGL} to do generic OpenGL state setup, as shown
in \fig{fWindowInit}.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/directSend.pdf}
  {\caption{\small\label{fDirectSend}Parallel Direct Send Compositing}}
\end{wrapfigure}
In direct send compositing, each rendering channel is also responsible
for the sort-last composition of one screen-space tile. He receives
the framebuffer pixels for his tile from all the other channels. The
size of one tile decreases linearly with the number of source channels,
which keeps the total amount of pixel data per channel constant.

@@

\fig{fDirectSend} illustrates this algorithm for three channels. The
Equalizer website contains a
presentation\footnote{\link{http://www.equalizergraphics.com/documents/EGPGV07.pdf}}
explaining and comparing this algorithm with the binary-swap algorithm.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/assembly.pdf}
  {\caption{\small\label{fAssembly}Hierarchy of assembly classes}}
\end{wrapfigure}
Readback and assemble operations on frames and images are designed to be
asynchronous. They have a start and finish method for both readback and
assemble to allow the initiation and synchronization of the operation.
Currently, only synchronous readback and assembly using
\textsf{glReadPixels} and \textsf{glDrawPixels} is implemented in the
respective start method of the image. Later versions of Equalizer will
implement asynchronous pixel transfers.

The offset of input and output frames characterizes the position of the
frame data with respect to the framebuffer, that is, the
\textbf{window's} lower-left corner. For output frames this is the
position of the channel with respect to the window.

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/volResult.png}
  {\caption{\small\label{fVolResult}Result of \fig{fBlendPerspective}}}
\end{wrapfigure}
DB compounds have the advantage of scaling any part of the volume
rendering pipeline: texture and main memory (smaller bricks for each
channel), fill rate (less samples per channel) and IO bandwidth for
time-dependent data (less data per time step and channel). Since the
amount of texture memory needed for each node decreases linearly, they
make it possible to render data sets which are not feasible to
visualize with any other approach.

@@

Finding the correct assembly order for perspective frustra is more
complex. The perspective distortion invalidates a simple orientation
criteria like the one used for orthographic frustra. For the view and
frustum setup shown in \fig{fBlendPerspective}\footnote{Volume Data Set
  courtesy of: AVS, USA} the correct compositing order is 4-3-1-2 or
1-4-3-2.

>>>>

\date{
  \vspace{2cm}
  \begin{figure}[ht]
    \hspace{-4.5cm}\includegraphics[width=22cm]{images/teaser.pdf}
  \end{figure}
  \vspace{1cm}
%  \textbf{RELEASE CANDIDATE}\\[\medskipamount]
  Version 1.3.7 for Equalizer 0.5.4+svn\\[\medskipamount]
  \today
}

@@

\clearpage
\thispagestyle{empty}
\tableofcontents
\vfill{\center\begin{tabularx}{\textwidth}{|l|l|X|}
    \hline
    \bf Rev & \bf Date     & \bf Changes \\
    \hline
    1.0     & Oct 28, 2007 & Initial Version for Equalizer 0.4\\
    1.2     & Apr 15, 2008 & Version for Equalizer 0.5\\
    1.3.1   & May 02, 2008 & Add \sref{sStatistics}, improved images based on 1.2 print\\
    1.3.2   & May 05, 2008 & Add network distribution on eqPly model data\\
    1.3.3   & May 10-19, 2008 & add \aref{aFileFormat}\\
    1.3.4   & July 3, 2008 & Update \sref{sStatistics}\\
    1.3.5   & July 7, 2008 & Update after namespace naming cleanup\\
    1.3.6   & July 20, 2008 & Add accum and samples to \aref{aFileFormat}, add \sref{sDrawableConfig}\\
    1.3.7   & Aug 22, 2008 & Update \sref{sEqplyWIndow} with OSWindow changes (event handling updates still missing)\\
    \hline
  \end{tabularx}}
\thispagestyle{empty}
\clearpage

@@

The \textsf{eq::Window} class encapsulates a drawable and an OpenGL
context. The drawable can be an on-screen window or an off-screen
PBuffer. Framebuffer object (FBO) support will be implement by a later
version of Equalizer. The window uses an \textsf{eq::OSWindow}, which
abstracts and manages window-system-specific handles to the drawable and
context, e.g., an X11 window \textsf{XID} and \textsf{GLXContext} for
the glX window system.

@@

\subsection{\label{sEqplyWIndow}Window}

The Equalizer window abstracts an OpenGL drawable and a rendering
context. When using the default window initialization functions, all
windows of a pipe share the OpenGL context. This allows reuse of OpenGL
objects such as display lists and textures between all windows of one
pipe.

@@

\subsubsection{Window System Interface}

The particulars of creating a window and OpenGL context depend on the
window system used. One can either use the implementation provided by
the operating system, e.g., AGL, WGL or glX, or some higher-level
toolkit, e.g., Qt.

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/osWindow.pdf}
  {\caption{\small\label{fOSWindow}eq::OSWindow UML class hierarchy}}
\end{wrapfigure}
All window-system specific functionality is implemented by a
specialization of \textsf{eq::OS\-Window}. The \textsf{OSWindow} class
defines the minimal interface to be implemented for a new window
system. Each \textsf{Window} uses one \textsf{OSWindow} during
execution. This separation allows an easy implementation and adaption to
another window system or application.

Equalizer provides a generic interface and implementation for the three
most common window systems: AGL, WGL and glX. The interfaces define the
minimal functionality needed to reuse other window system specific
classes, for example the AGL, WGL and glX event handlers. The
implementation derived from these interfaces provides a sample
implementation which honors all configurable window attributes.

@@

The initialization sequence uses multiple, overrideable task
methods. The main task method \textsf{configInit} calls first
\textsf{configInitOSWindow}, which creates and initializes the
\textsf{OSWindow} for this window. The \textsf{OSWindow} initialization
code is implementation specific. If the \textsf{OSWindow} was
initialized successfully, \textsf{configInit} calls
\textsf{configInitGL}, which performs the generic OpenGL state
initialization. The default implementation sets up some typical OpenGL
state, e.g., it enables the depth test.

The \textsf{OSWindow} initialization takes into account various
attributes set in the configuration file. Attributes include the size of
the various frame buffer planes (color, alpha, depth, stencil) as well
as other framebuffer attributes, such as quad-buffered stereo,
doublebuffering, fullscreen mode and window decorations. Some of the
attributes, such as stereo, doublebuffer and stencil can be set to
\textsf{eq::AUTO}, in which case the Equalizer default implementation
will test for their availability and enable them if possible.

For the window-system specific initialization, \textsf{eqPly} uses the
default Equalizer implementation. The \textsf{eqPly} window
initialization only overrides the OpenGL-specific initialization
function \textsf{configInitGL} in order to initialize a state object and
an overlay logo. This function is only called if an OpenGL context was
created and made current:

@@

    glLightModeli( GL_LIGHT_MODEL_LOCAL_VIEWER, 1 );
    glEnable( GL_CULL_FACE ); // OPT - produces sparser images in DB mode
    glCullFace( GL_BACK );

    EQASSERT( !_state );
    _state = new VertexBufferState( getObjectManager( ));
    _loadLogo();

    const Config*   config   = static_cast< const Config* >( getConfig( ));
    const InitData& initData = config->getInitData();
    if( initData.useGLSL() )
        _loadShaders();

@@

The state object is used to handle the creation of OpenGL objects in a
multipipe, multi-threaded execution environment. It uses the object
manager of the \textsf{eq::Window}, which is described in detail in
\sref{sObjectManager}.

@@

The window exit happens in the reverse order of the
initialization. First, \textsf{configExitGL} is called to de-initialize
OpenGL, followed by \textsf{configExitOSWindow} which de-initializes the
drawable and context and deletes the \textsf{OSWindow} allocated in
\textsf{configInitOSWindow}.

The window OpenGL exit function of \textsf{eqPly} de-allocates all
OpenGL objects. The object manager does not delete the object in its
destructor, since it does not know if an OpenGL context is still
current.

{\footnotesize\begin{lstlisting}
bool Window::configExitGL()
{
    if( _state )
        _state->deleteAll();

    delete _state;
    _state = 0;

    return eq::Window::configExitGL();
}
\end{lstlisting}}

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/render.pdf}
  {\caption{\small\label{fRender}Main Render Loop}}
\end{wrapfigure}
The model data is spatially organized in an 3-dimensional
kD-tree\footnote{See also \link{http://en.wikipedia.org/wiki/Kd-tree}}
for efficient view frustum culling. When the model is loaded by
\textsf{Node::config\-Init}, it is preprocessed into the kD-tree and each
node of the tree gets a database range assigned. The root node
has the range [0, 1], its left child [0, 0.5] and its right child [0.5,
1], and so on for all nodes in the tree. The preprocessed model is saved
in a binary format for accelerating subsequent use.

The rendering loop maintains a list of candidates to render, which
initially contains the root node. Each candidate of this list is tested
for full visiblity against the frustum and range, and rendered if
visible. It is dropped if it is fully invisible or fully out of
range. If it is partially visible or partially in range, the children of
the node are added to the candidate list. \fig{fRender} shows a flow
chart of the rendering algorithm, which performs efficient view frustum
and range culling.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/threads.pdf}
  {\caption{\small\label{fThreads}Threads within one node process}}
\end{wrapfigure}
During this initialization, Equalizer creates and manages two threads
for communication, the receiver thread and the command thread. Normally
no application code is executed from these two threads.

@@

The command thread processes Equalizer-related request from other nodes,
for example during \textsf{eq::net::Object} mapping. In some special cases
the command thread executes application code, for example when a remote
node maps a static or unbuffered object,
\textsf{Object::getInstanceData} is called from the command thread.

@@

Each entity uses \textsf{waitFrame\-Started} to block on the parent's
\textsf{startFrame}, e.g., \textsf{Pipe::\-frame\-Start} calls
\textsf{Node::wait\-Frame\-Started} to wait for the corresponding
\textsf{Node::start\-Fra\-me}. This explicit synchronization allows to
update non-critical data before synchronizing with
\textsf{waitFrameStarted}, or after unlocking using
\textsf{start\-Fra\-me}. \fig{fFrameSync} illustrates this
synchronization model.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/frameSync.pdf}
  {\caption{\small\label{fFrameSync}Per-Node Frame Synchronization}}
\end{wrapfigure}
\fig{fFrameSync} illustrates the local frame synchronization. By
removing the calls to \textsf{waitFrameLocal}, the node thread can
disable the synchronization with the pipe threads, which causes the node
operations to overlap with the rendering.

@@

This section explains window initialization in detail. It discusses in
detail the handling on the different window systems. The entry point for
the window initialization is the task method
\textsf{Window::configInit}. This task method first calls a
window-system-specific task method, and then
\textsf{Window::configInitGL} to do generic OpenGL state setup.

@@

In direct send compositing, each rendering channel is also responsible
for the sort-last composition of one screen-space tile. He receives
the framebuffer pixels for his tile from all the other channels. The
size of one tile decreases linearly with the number of source channels,
which keeps the total amount of pixel data per channel constant.

@@

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/directSend.pdf}
  {\caption{\small\label{fDirectSend}Parallel Direct Send Compositing}}
\end{wrapfigure}
\fig{fDirectSend} illustrates this algorithm for three channels. The
Equalizer website contains a
presentation\footnote{\link{http://www.equalizergraphics.com/documents/EGPGV07.pdf}}
explaining and comparing this algorithm with the binary-swap algorithm.

@@

Readback and assemble operations on frames and images are designed to be
asynchronous. They have a start and finish method for both readback and
assemble to allow the initiation and synchronization of the operation.
Currently, only synchronous readback and assembly using
\textsf{glReadPixels} and \textsf{glDrawPixels} is implemented in the
respective start method of the image. Later versions of Equalizer will
implement asynchronous pixel transfers.

\begin{wrapfigure}{r}{.618\textwidth}
  \includegraphics[width=.618\textwidth]{images/assembly.pdf}
  {\caption{\small\label{fAssembly}Hierarchy of assembly classes}}
\end{wrapfigure}
The offset of input and output frames characterizes the position of the
frame data with respect to the framebuffer, that is, the
\textbf{window's} lower-left corner. For output frames this is the
position of the channel with respect to the window.

@@

DB compounds have the advantage of scaling any part of the volume
rendering pipeline: texture and main memory (smaller bricks for each
channel), fill rate (less samples per channel) and IO bandwidth for
time-dependent data (less data per time step and channel). Since the
amount of texture memory needed for each node decreases linearly, they
make it possible to render data sets which are not feasible to
visualize with any other approach.

@@

\begin{wrapfigure}{r}{.382\textwidth}
  \includegraphics[width=.382\textwidth]{images/volResult.png}
  {\caption{\small\label{fVolResult}Result of \fig{fBlendPerspective}}}
\end{wrapfigure}
Finding the correct assembly order for perspective frustra is more
complex. The perspective distortion invalidates a simple orientation
criteria like the one used for orthographic frustra. For the view and
frustum setup shown in \fig{fBlendPerspective}\footnote{Volume Data Set
  courtesy of: AVS, USA} the correct compositing order is 4-3-1-2 or
1-4-3-2.
